---
title: "What to Expect with Life Expectancy?"
author: 
  - Karmen Zhu, Edan Wong, Saanvi Prasanth
date: 4 October 2024
date-format: long
format: pdf
number-sections: true
---

```{r}
library(tidyverse)
data <- LifeExpectancyData

colnames(data)[colnames(data) == "Life expectancy"] <- "LifeExpectancy"
colnames(data)[colnames(data) == "Income composition of resources"] <- "IncomeComposition"

data <- data[, c("LifeExpectancy", "GDP", "Polio", "Diphtheria", "Status", "IncomeComposition")]
# View the selected columns
head(selected_columns)
#### Save data ####
write_csv(cleaned_data, "~/STA302-group55/LifeExpectancy/LifeExpectancy/data/raw_data/cleaned_data.csv")
```

# Data {#sec-data}

```{r}
# Randomly select 10 rows from the data
sampled_data <- data |> sample_n(10)

# Display the selected data
sampled_data
```


# Results {#sec-mydatasection}

## Response variable(Life Expectancy)
```{r}

morandi_hist <- rgb(0.2, 0.0, 0.0, 0.4)
  hist(data$LifeExpectancy,
     main = "Distribution of Life Expectancy",
     xlab = "Life Expectancy",
     col = morandi_hist,
     breaks = 10)
```


# Visualizing the Relationship Between Life Expectancy and Its Predictors
```{r}

par(mfrow = c(2, 3))

# 1. Life Expectancy VS GDP plot
morandi_color_gdp <- rgb(0.4, 0.6, 0.4, 0.4)
plot(data$GDP, data$LifeExpectancy, 
     main = "Life Expectancy VS GDP", 
     xlab = "GDP", 
     ylab = "Life Expectancy",
     col = morandi_color_gdp,
     pch = 19,
     cex = 0.5)

# 2. Life Expectancy VS Polio Immunization Coverage plot
morandi_color_polio <- rgb(0.8, 0.6, 0.6, 0.4)
plot(data$Polio, data$LifeExpectancy, 
     main = "VS Polio Immunization Coverage", 
     xlab = "Polio", 
     ylab = "Life Expectancy",
     col = morandi_color_polio,
     pch = 19,
     cex = 0.5)

# 3. Life Expectancy VS Diphtheria Immunization Coverage plot
morandi_color_diphtheria <- rgb(0.6, 0.8, 0.8, 0.4)
plot(data$Diphtheria, data$LifeExpectancy, 
     main = "VS Diphtheria Immunization Coverage", 
     xlab = "Diphtheria", 
     ylab = "Life Expectancy",
     col = morandi_color_diphtheria,
     pch = 19,
     cex = 0.5)

# 4. Linear Model for Life Expectancy by Developed/Developing Status
data$Status <- factor(data$Status, labels = c("Developed", "Developing"))
boxplot(LifeExpectancy ~ Status, data=data, 
        main="VS Status", 
        xlab="Status", ylab="Life Expectancy")

# 5. Life Expectancy VS Income Composition of Resources plot
morandi_color_income <- rgb(0.6, 0.6, 0.8, 0.4)
plot(data$IncomeComposition, data$LifeExpectancy, 
     main = "VS Income Composition of Resources", 
     xlab = "Income Composition of Resources", 
     ylab = "Life Expectancy",
     col = morandi_color_income,
     pch = 19,
     cex = 0.5)

# Reset layout back to normal (1 plot per page)
par(mfrow = c(1, 1))

```


# Preliminary Model using the Predictors
```{r}

fit = lm(LifeExpectancy ~  GDP + Polio + Diphtheria + Status + IncomeComposition, data=data)
summary(fit)
```

# Correlation Matrix
```{r}

# Convert Status to a numeric variable
data$Status <- as.numeric(factor(data$Status))
cor_data <- data[, c("LifeExpectancy", "GDP", "Polio", "Diphtheria", "IncomeComposition")]
# correlation matrix
cor_matrix <- cor(cor_data, use = "complete.obs")
print(cor_matrix)
```
# Correlation Plot
```{r}

library(PerformanceAnalytics)
library(dplyr)
cor_data <- data[, c("LifeExpectancy", "GDP", "Polio", "Diphtheria", "IncomeComposition")]
chart.Correlation(cor_data, histogram = TRUE, pch = 19)
```

# Residuals vs Fitted
```{r}

morandi_residual_fitted <- rgb(1.0, 0.8, 0.4, 0.4)
fi <- fitted(fit)
resi <- resid(fit)
plot(resi ~ fi,
     main = "Residuals vs Fitted",
     xlab = "Fitted values",
     ylab = "Residuals",
     pch = 19, col = morandi_residual_fitted)
abline(h = 0, col = "red", lty = 2)
```


```{r}
#| label: fig-residuals-plots
#| fig-cap: Residuals versus Each Predictors
#| echo: false

library(car)
fitted <- lm(LifeExpectancy ~ GDP + Polio + Diphtheria + IncomeComposition, data = data)
crPlots(fitted, smooth = list(span = 0.8))  # 'span' is for stability
```


# Q-Q Plot of Residuals
```{r}

morandi_qq <- rgb(0.4, 0.2, 0.2, 0.4)
qqnorm(resi, main = "Q-Q Plot", col = morandi_qq)
qqline(resi, col = "red", lty = 2)
```

# A root of Standardized Residuals vs Fitted Values
```{r}

morandi_scale <- rgb(0.0, 0.4, 1.0, 0.4)
standardized_residuals <- rstandard(fit)
plot(fi, sqrt(abs(standardized_residuals)),
     main = "Scale-Location",
     xlab = "Fitted Values",
     ylab = "√|Standardized Residuals|",
     pch = 19, col = morandi_scale)

# Add a smoothing line to check for patterns
lines(lowess(fi, sqrt(abs(standardized_residuals))), col = 'red', lwd = 2)
```


# Standardized Residuals vs Leverage
```{r}

morandi_leverag <- rgb(0.6, 0.2, 0.0, 0.4)
plot(fit, which = 5, main = "Residuals vs Leverage Plot", col = morandi_leverag)


leverage_values <- hatvalues(fit)

# Display leverage values
leverage_values

# Set a threshold for high leverage points (commonly 2 times the average leverage value)
threshold <- 2 * mean(leverage_values)
# Identify and display observations with high leverage
high_leverage_points <- which(leverage_values > threshold)
high_leverage_values <- leverage_values[high_leverage_points]
```

# Histogram of Residuals (to check the distribution of residuals)
```{r}

morandi_hist <- rgb(0.2, 0.4, 0.6, 0.4)
hist(resi, breaks = 20, main = "Histogram of Residuals", 
     xlab = "Residuals", col = morandi_hist, border = "white")
```
# Cook's distance
```{r}
cooksd <- cooks.distance(fit)

# Generate Cook's Distance plot
morandi_cook <- rgb(0.6, 0.1, 0.2, 0.6)
plot(cooksd, type = "h", lwd = 2, col = morandi_cook, 
     main = "Cook's Distance", ylab = "Cook's Distance")

# Add a threshold line for influential points (commonly 4/n where n is the number of observations)
abline(h = 4 / nrow(data), col = "red", lty = 2)
```



